Data Provenance Tracking for Concurrent Programs

### 2014-03-20 paper.tex
committing the paper with this new figure in the intro

\newcommand{\Caption}[1]{\begin{minipage}{1.0\columnwidth} \caption{#1} \end{minipage}}
\newcommand{\CaptionWide}[1]{\caption{#1}\vspace{-2ex}}
%\doublespacing
\authorinfo{Anonymous for Submission}{}

@@

Our approach to solving that problem is to provide the programmer with {\em
provenance} information that can be examined alongside the program's memory
state ({\em i.e.}, core dump).  A value's provenance describes why a memory
location contains its value.  In this work we develop a new provenance tracking
mechanism called {\em last writer slicing} that satisfies these goals.  A
memory location's last writer slice is a record containing the program point
and the identifier of the thread that last wrote data to that memory location.
As the program executes in production, we collect last writer slices for each
potentially shared memory location and those last writer slices record the
provenance of the values stored in those locations.  Last writer slices are
saved with a core dump when the program fails and are available in the debugger
({\em e.g.}, for breakpoint debugging).

@@

Motivated by these prior {\em ad hoc} efforts, we build a general mechanism
called {\em Communication Traps} or {\em \ctraps} on top of last writer slices
(Section~\ref{sec:ctraps}). \ctraps is useful for implementing dynamic
concurrency analyses, like the ones discussed above.  \ctraps runs alongside a
program and uses last writer slices to identify memory accesses that lead to
{\em communication} between the threads. Communcation occurs when a thread
accesses memory last written by a different thread and \ctraps delivers a {\em
trap} to a thread when it communicates with another thread, giving it
information about how communication occurred.  Programmer-defined {\em trap
handlers} can implement arbitrary analysis using that information. By default,
\ctraps has low overheads that are appropriate production in many cases.  As we
show in our evaluation, the overhead of \ctraps applications scales with the
complexity of their analysis.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/JDKStringBufferFail.pdf}
\caption{\label{fig:coreDumpFail}{\bf Atomicity violation bug in a string buffer implementation.} The core dump does not reveal the connection between {\tt append()} and {\tt erase()}, which is crucial to understand the bug.}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=.6]{figs/BasicLWT.pdf}
\caption{\label{fig:basicLWT}Basic operation of the last writer table. Different shapes signify different distinct memory operations. }
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/JDKStringBufferDebug2.pdf}
\caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the buggy interaction between {\tt erase()} and {\tt append()}.}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=.9\columnwidth]{figs/LWS_Overview.pdf}
\caption{\label{fig:lwsoverview}{\bf Overview of system support for last writer slicing.} Our compiler and runtime library ensure the \lwt remains updated so it can be used during debugging.}
\end{figure}

@@

\paragraph{Non-overlapping Memory References} 
We assume written memory locations do not overlap in memory because our \lwt
design is {\em granularity oblivious}.  We discuss this issue informally due to
space constraints.  Each \lwt entry corresponds to a location that was written.
Two instructions might reference different locations that overlap because the
referenced locations have different granularities ({\em e.g.}, accessing a byte
that is part of a previously accessed word).  When that happens, both accesses
update different \lwt entries, because they reference different locations.
However, the overlapping region then has two different entries reflecting the
last {\em two} updates to its parts rather than just its last write.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=.90\columnwidth]{figs/CTraps_Overview.pdf}
\caption{\label{fig:systemdiagram}{\bf CTraps system support.}}
\end{figure}

@@

\paragraph{Compiler Support}
Our compiler inserts a call to the \ctraps~{\em trap hook} function immediately
before all read and write operations that access potentially escaping memory
locations.  Trap hooks are passed the address of the memory location being
accessed, the thread identifier of the accessing thread, the type of operation
being performed, and the program point of the access.

@@

%\begin{figure}[htb]
%\centering
%\begin{verbatim}
%void trap(AccessType type, 
%          ThdId currThd, ProgPt currPt,
%          ThdId lastThd, ProgPt lastPt);
%\end{verbatim}
%\caption{\label{fig:hookapi}The \ctraps trap handling interface.}
%\end{figure}

@@

\begin{table}
\scriptsize
\centering
\begin{tabular}{l|ll|l|ccc|c}
{\bf App} & {\bf Ver. \#} & {\bf Bug \#} &{\bf Type} & {\bf Eff.?} & {\bf Triv.?} & {\bf Usef.?} & {\bf O.T.?}\\ \hline
jdk1.4 &  1.70   &  N/A     & Atom.  &Yes  &No    &Yes   & No  \\%All values usable
pbzip2 &  0.9.4  &  N/A     & Order  &Yes  &No    &Yes   & Yes \\
httrack& 3.43.9  &  N/A     & Order  &Yes  &No    &Yes   & No  \\%Used w/o origin - no help
transmission & 1.42 & N/A   & Order  &Yes  &No    &Yes   & No  \\
mysql  &  4.0.12 &  791     & Atom.  &Yes  &No    &Yes   & No  \\%All values usable
apache1&  2.0.48 &  21285   & Atom.  &Yes  &No    &Yes   & Some \\%All values usable
apache2&  2.2.9 &  45605    & Atom.  &Yes  &No    &Some  & No \\%All values usable
\end{tabular}
\caption{\label{tab:bugs}{\bf Concurrency debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.  }
\normalsize
\end{table}

@@

\paragraph{httrack}
Figure~\ref{fig:httlws} shows how last writer slices help understand the root
cause of a concurrency bug from HTTrack-3.43.9.  The program crashes at point
{\bf B} when it calls {\tt hts\_mutex\_lock()} with an invalid mutex.    The
root cause of this bug is that Thread 1 fails to initialize the mutex before
Thread 2 uses it.   The core dump is of little debugging use, only showing
that, indeed, the lock is uninitialized.  Even if the programmer compared the
core dump from the failing run to the memory image from a correct execution --
a helpful debugging strategy -- they would find only that in a correct
execution, the lock contains valid mutex state.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
ordering violation in HTTrack.} The provenance information stored in the last
writer slice for the mutex variable helps understand the bug's root cause.}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/MySQLDebug.pdf}
\caption{\label{fig:mysqllws}{\bf Using last writer slices to debug an
atomicity violation in MySQL 4.0.12.} The last writer slice reveals the code that closes the log before a database insertion that goes unlogged, which is a security problem.}
\end{figure}

@@

\paragraph{LevelDB-1.5.0}
LevelDB is a carefully tuned high-performance key-value store written by Google
developers. To benchmark LevelDB, we used the included {\tt db\_bench} utility,
running its ``Read while Writing'' test with 8 threads.  We measure LevelDB's
performance to be the operation throughput reported by {\tt db\_bench}.

@@

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{plots/perf.pdf}
\caption{\label{fig:perfall}Runtime overhead of \ctraps and Last Writer Slices (LWS).}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{plots/appperf.pdf}
\caption{\label{fig:perfapps}Performance overheads of CCI-Prev and Graph Collection implemented with \ctraps.}
\end{figure}

@@

Apache         &  21.4\%&(12K)                    \\
MySQL          &  21.3\%&(53K)                    \\
memcached      &  65.8\%&(7.5K)                   \\
LevelDB        &  44.0\%&(4.5K)                   \\ \hline
blackscholes   &   7.5\%&(4)                      \\
dedup          &   22.0\%&(171)                   \\
canneal        &   28.6\%&(226)                   \\
x264           &   29.8\%&(5960)                  \\
vips           &   24.6\%&(9116)                  \\
ferret         &   20.9\%&(792)                   \\
fluidanimate   &   19.6\%&(149)                   \\
streamcluster  &   11.6\%&(47)                    \\
swaptions      &   8.3\% &(19)                    \\
\end{tabular}
\caption{\label{tab:char}Instrumentation pruned by static analysis.}
\end{table}

>>>>

\newcommand{\Caption}[1]{\begin{minipage}{.95\columnwidth} \caption{#1} \end{minipage} \vspace{-1.2ex}}
\newcommand{\CaptionWide}[1]{\caption{#1}\vspace{-2ex}}
%\doublespacing
\authorinfo{Anonymous for Submission}{}

@@

Our approach to solving that problem is to provide the programmer with {\em
provenance} information that can be examined alongside the program's memory
state ({\em i.e.}, core dump).  A value's provenance describes why a memory
location contains its value.  In this work we develop a new provenance tracking
mechanism called {\em last writer slicing}.  A
memory location's last writer slice is a record containing the program point
and the identifier of the thread that last wrote data to that memory location.
As the program executes in production, we collect last writer slices for each
potentially shared memory location and those last writer slices record the
provenance of the values stored in those locations.  Last writer slices are
saved with a core dump when the program fails and are available in the debugger
({\em e.g.}, for breakpoint debugging).

@@

Motivated by these prior {\em ad hoc} efforts, we develop a dynamic analysis
framework called {\em Communication Traps} or {\em \ctraps}.  \ctraps runs
alongside a program and uses the provenance information provided by last writer
slices to identify memory accesses that lead to {\em communication} between the
threads. Communcation occurs when a thread accesses memory last written by a
different thread and \ctraps delivers a {\em trap} to a thread when it
communicates with another thread, giving it information about how communication
occurred.  Programmer-defined {\em trap handlers} can implement arbitrary
analysis using that information. By default, \ctraps has low overheads that are
appropriate production in many cases.  As we show in our evaluation, the
overhead of \ctraps applications scales with the complexity of their analysis.
Figure~\ref{fig:overview} illustrates the relationship between last writer
slices, \ctraps, and \ctraps applications.

\begin{figure}[h]
\centering
\includegraphics[width=.70\columnwidth]{figs/overview.pdf}
\Caption{\label{fig:overview}{\bf An overview of last writer slicing and CTraps.} Last writer slicing is a basic mechanism for tracking provenance.  The CTraps framework uses last writer slicing to identify communication and expose it to CTraps applications.  CTraps applications implement arbitrary dynamic concurrency analyses, like communication graph collection~\cite{recon} or cooperative debugging~\cite{cci}.}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/JDKStringBufferFail.pdf}
\Caption{\label{fig:coreDumpFail}{\bf Atomicity violation bug in a string buffer implementation.} The core dump does not reveal the connection between {\tt append()} and {\tt erase()}, which is crucial to understand the bug.}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=.6]{figs/BasicLWT.pdf}
\Caption{\label{fig:basicLWT}Basic operation of the last writer table. Different shapes signify different distinct memory operations. }
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/JDKStringBufferDebug2.pdf}
\Caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the buggy interaction between {\tt erase()} and {\tt append()}.}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=.9\columnwidth]{figs/LWS_Overview.pdf}
\Caption{\label{fig:lwsoverview}{\bf Overview of system support for last writer slicing.} Our compiler and runtime library ensure the \lwt remains updated so it can be used during debugging.}
\end{figure}

@@

\paragraph{Non-overlapping Memory References} 
We assume written memory locations do not overlap in memory because our \lwt
design is {\em granularity oblivious}.  Each \lwt entry corresponds to a
location that was written.  Two instructions might reference different
locations that overlap because the referenced locations have different
granularities ({\em e.g.}, accessing a byte that is part of a previously
accessed word).  When that happens, both accesses update different \lwt
entries, because they reference different locations.  However, the overlapping
region then has two different entries reflecting the last {\em two} updates to
its parts rather than just its last write.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=.90\columnwidth]{figs/CTraps_Overview.pdf}
\Caption{\label{fig:systemdiagram}{\bf CTraps system support.}}
\end{figure}

@@

\paragraph{Compiler Support}
Our compiler inserts a call to the CTraps {\em trap hook} function immediately
before all read and write operations that access potentially escaping memory
locations.  Trap hooks are passed the address of the memory location being
accessed, the thread identifier of the accessing thread, the type of operation
being performed, and the program point of the access.

@@

%\begin{figure}[htb]
%\centering
%\begin{verbatim}
%void trap(AccessType type, 
%          ThdId currThd, ProgPt currPt,
%          ThdId lastThd, ProgPt lastPt);
%\end{verbatim}
%\Caption{\label{fig:hookapi}The \ctraps trap handling interface.}
%\end{figure}

@@

\begin{table}
\scriptsize
\centering
\begin{tabular}{l|ll|l|ccc|c}
{\bf App} & {\bf Ver. \#} & {\bf Bug \#} &{\bf Type} & {\bf Eff.?} & {\bf Triv.?} & {\bf Usef.?} & {\bf O.T.?}\\ \hline
jdk1.4 &  1.70   &  N/A     & Atom.  &Yes  &No    &Yes   & No  \\%All values usable
pbzip2 &  0.9.4  &  N/A     & Order  &Yes  &No    &Yes   & Yes \\
httrack& 3.43.9  &  N/A     & Order  &Yes  &No    &Yes   & No  \\%Used w/o origin - no help
transmission & 1.42 & N/A   & Order  &Yes  &No    &Yes   & No  \\
mysql  &  4.0.12 &  791     & Atom.  &Yes  &No    &Yes   & No  \\%All values usable
apache1&  2.0.48 &  21285   & Atom.  &Yes  &No    &Yes   & Some \\%All values usable
apache2&  2.2.9 &  45605    & Atom.  &Yes  &No    &Some  & No \\%All values usable
\end{tabular}
\Caption{\label{tab:bugs}{\bf Concurrency debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.  }
\normalsize
\end{table}

@@

\paragraph{httrack}
Figure~\ref{fig:httlws} shows how last writer slices help understand the root
cause of a concurrency bug from HTTrack-3.43.9.  The program crashes at point
{\bf B} when it calls {\tt hts\_} {\tt mutex\_lock()} with an invalid mutex.    The
root cause of this bug is that Thread 1 fails to initialize the mutex before
Thread 2 uses it.   The core dump is of little debugging use, only showing
that, indeed, the lock is uninitialized.  Even if the programmer compared the
core dump from the failing run to the memory image from a correct execution --
a helpful debugging strategy -- they would find only that in a correct
execution, the lock contains valid mutex state.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
\Caption{\label{fig:httlws}{\bf Using last writer slices to debug an
ordering violation in HTTrack.} The provenance information stored in the last
writer slice for the mutex variable helps understand the bug's root cause.}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/MySQLDebug.pdf}
\Caption{\label{fig:mysqllws}{\bf Using last writer slices to debug an
atomicity violation in MySQL 4.0.12.} The last writer slice reveals the code that closes the log before a database insertion that goes unlogged, which is a security problem.}
\end{figure}

@@

\paragraph{LevelDB-1.5.0}
LevelDB is a high-performance key-value store written by Google
developers. To benchmark LevelDB, we used the included {\tt db\_bench} utility,
running its ``Read while Writing'' test with 8 threads.  We measure LevelDB's
performance to be the operation throughput reported by {\tt db\_bench}.

@@

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{plots/perf.pdf}
\Caption{\label{fig:perfall}Runtime overhead of \ctraps and Last Writer Slices (LWS).}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{plots/appperf.pdf}
\Caption{\label{fig:perfapps}Performance overheads of CCI-Prev and Graph Collection implemented with \ctraps.}
\end{figure}

@@

Apache         &  21.4\%&(12K)                    \\
MySQL          &  21.3\%&(53K)                    \\
memcached      &  65.8\%&(7.5K)                   \\
LevelDB        &  44.0\%&(4.5K)                   \\ \hline
blackscholes   &   7.5\%&(4)                      \\
dedup          &   22.0\%&(171)                   \\
canneal        &   28.6\%&(226)                   \\
x264           &   29.8\%&(5960)                  \\
vips           &   24.6\%&(9116)                  \\
ferret         &   20.9\%&(792)                   \\
fluidanimate   &   19.6\%&(149)                   \\
streamcluster  &   11.6\%&(47)                    \\
swaptions      &   8.3\% &(19)                    \\
\end{tabular}
\Caption{\label{tab:char}Instrumentation pruned by static analysis.}
\end{table}
